---
title: "Work Life Balance"
author: "Anup"
date: "06/10/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE}
library(readxl)
library(haven)
library(finalfit)
library(dplyr)
library(ggplot2)
library(likert)
library(patchwork)
library(knitr)
library(kableExtra)
library(tidyverse)
library(tidytext)
```

### Motivation of the Study

Health makes an important contribution to economic progress, as healthy
population live longer, is more productive, and save more. This has huge
human and economic costs: India is losing more than six percent of its
GDP annually due to premature deaths and preventable illnesses (WHO,
2013). The Indian Health Care sector is prominently the largest and
fastest growing sector in the country and it comprises of Hospital
(71%)\*, Pharmaceutical (13%), Diagnosis (3%), Medical equipments and
supplies (9%) and medical insurance (4%). In India, as per information
provided by Medical Council of India (MCI) and Indian Nursing Council,
there are 9,38,861 Doctors, 7,86,061 ANMs and 17,80,006 Nurses &
Midwives registered in the country. The shortage of qualified medical
professionals is one of the key challenges the Indian healthcare
industry is facing (Source: WHO, IMH, Deloitte, Tech Sci. Research). The
health care sector totally relies in the hand of health care providers
and undoubtedly, doctors and nurses are the dominant personnel in health
care. Job stress among healthcare staff is becoming a common occurrence
in most public health services (Winstanley and Whittington, 2002).

The daily work of physicians and nurses in the hospital setting is
inherently stressful (Firth-Cozens, 2003; Lee & Wang, 2002; Thomas,
2004). Long workdays, excess case loads, time pressure, poor sleep
habit, and high performance expectation contribute to the
well-documented rates of burnout, depression, job dissatisfaction, and
workplace fatigue among nurses and physicians (Clever,2002; Geurts,
Rutte, & Peeters, 1999; Hillhouse & Adler, 1997; Imai, Nakao, Tsuchiya,
Kuroda, & Katoh, 2004; Papp, Stoller, Sage, Aikens, Owens, Avidan et
al., 2004; Rose, Ware, Kolm, & Risser, 2000; Shanafelt, Bradley, Wipf, &
Back, 2002). The consequences of these characteristics may include
decreased patient satisfaction, diminished care quality, and increased
rate of medical errors (DiMatteo, Sherbourne, Hays, Ordway, Kravitz,
McGlynn et al., 1993; Fletcher, Davis, Underwood, Mangrulkar, McMahon, &
Saint, 2004 Grol, Mokkink, Smits, van Eijk, Beek, Mesker et al., 1985;
Tarnow- Mordi, Hau, Warden, & Shearer, 2000; Weinger & Ebden, 2002).
Medical practitioners experience appreciable stress (Burke and
Richardson, 1991), comparatively high suicide rates (Gestal, 1987), and
varying degree of morbidity and early retirement (Mc Namee, Keen,
Corkill, 1987). The main sources of stress that have been identified
are: problems with practice administration, interruptions, patient's
expectations and demands, emergencies, heavy workload, constant time
pressure and work/home conflict (Cooper et al., 1989; Howie et al.,
1989; Morrell et al., 1986; Porter et al., 1985; Richardson & Burke,
1991). Health Care Personnel are often stressed with their own work
which some where affects their family life in an adverse manner.
According to "The Ottawa Citizen, 2008", in about 2,800 medical students
and 730 second-year residents, about 93 per cent of students and 88 per
cent of residents said the work-life balance was an important part of
achieving a satisfying and successful medical practice. In the second
study, the Australian Medical Association (2008) revealed that 81 per
cent of hospital doctors surveyed want flexible working arrangements to
allow them to spend more time with family and to continue further formal
training. Hence, the concept of work-life balance is going to be very
appropriate for the healthcare professionals who are tend to work 24\*7
and involve a lot strain and stress.

## Research Gap

After literature review it was found that there are many studies
conducted in various fields and disciplines but there is a huge research
gap on the study of work-life balance in health care sector in Indian
perspective. Few studies has been carried out on this issue in Indian
Health Care Sector, either focusing on nurses or doctors but not both
together. Hence, this is the peak time to take an in depth understanding
on these issues which are not only desirable but also necessary. The
present research tries to find out causes of imbalances among doctors
and nurses. The results of the present study would be very useful to
government and health care sector to understand the causes of distress
and imbalance among the health care personnel and to make policies and
strategies so that they can work with making an optimum blend of work
and non-work domain, so that none of the domain prevails over the other.

## A Conceptual Framework of Work-Life Balance

"Work" and "Life" have been defined differently in literature
(Guest,2002), where work is paid employment and life is everything
outside of the formal employment but is usually used to connate the
sphere of family or home life (Ransome,2007). WLB is defined as
"satisfaction and good functioning at work and at home with a minimum of
role conflict" (Clark, 2000). WLB is defined as "the absence of
unacceptable level of conflicts between work and non-work demands.
(Greenbatt,2002) Work life balance as "the extent to which an individual
is equally engaged in -- and equally satisfied with -- his or her work
role and family role" (Greenhaus et al, 2003). Work life balance is "the
extent to which an individual's effectiveness and satisfaction in work
and family roles are compatible with the individual's life priorities.
(Greenhaus and Allen, 2006) Work--family balance is defined "as
accomplishment of role related expectations that are negotiated and
shared between an individual and his or her role-related partners in the
work and family domains" (Grzywacz and Carlson, 2007). WLB for any
person is having the "right" combination of participation in paid work
(defined by hours and working conditions) and other aspects of lives.
This combination will change as people move through life and have
changing responsibilities and commitments in their work and personal
lives. Frone, Russell, & Cooper, (1992). With these definitions it is
clear that any imbalance that is formed either creates more pressure or
psychological involvement or satisfaction towards one role as compared
to other role. The origin of the research work life balance can be seen
from the seminal work of rapport and rapport (1969), the research are
been that both work and family needs time and energy. Work is an
important source of income, financial stability and status. As work and
family has no direct relationship with each other so conflicts is
inevitable to arise. Khan wolf, Quinn, snoek and Rosenthal, (1964) from
their seminal studies associated work life balance with role theory,
that is conflicting expectations associated with different roles which
the male and female has to play in their day to day life. Kanter (1977)
was able to continue the research of Khan (1964) proving that work and
family are not independent of each other rather have an interlink which
consequently brings in conflicts between the two. With the pioneering
work of Pleck (1977) there was a general consensus formed that work and
family affect each other either in a positive or negative way. And there
are various extraneous variables like time, task, attitudes, stress
emotions, and behaviors which have an influence over work life
imbalance.

    Carlson and Frone (2003) discussed about the psychological involvement of the person in the organization as well as on the family front. A high degree of psychological involvement with context to role of an individual i.e. job role and family role would prioritize the individual to be somewhat mentally preoccupied with their performance in that role, such imbalance in performance will have significant implications on work to family and family to work interference. Conceptually, it seems evident that psychological involvement would have significant implications for levels of work to family and family to work interference. Carlson and Frone (2003) found that both psychological involvement and behavioral (time) involvement were significant related to WIF but the reverse did not apply with FIW or family involvement into work. Clearly more research and dynamic interplay in relationship of time and family was required to draw conclusions. Therefore, Frone (2003) summarized the findings from US based survey and noted that family boundaries may be more permeable than job boundaries and hence the levels of work to family (WFI) interference are reported higher or more intense than those of family to work interference (FWI).

### OBJECTIVES OF THE STUDY

The objectives framed for the study are as follows: 1. To explore the
variations in work-life balance indicators in the Indian and Global
context; 2. To identify the important variables related to work-life
balance in Health Care Sector (Hospitals); 3. To measure the dependency
of work-family conflict and family-work conflict on control variables;
4. To examine the existing work-life balance variables among doctors and
nurses; and 5. To assess the impact of work-life balance variables on
job satisfaction, employee well- being and stress control.

### HYPOTHESES FORMULATION

On the basis of literature review and above objectives the following
hypotheses are laid down: H01: There is no variation in work-life
balance variables in the Indian and Global context. H02: There is no
dependency of work-family conflict and family-work conflict on control
variables; H03: There is no significant relationship of work-life
balance variables among doctors and nurses; and H04: Work-Life Balance
variables do not significantly impact the job satisfaction, employee
well-being and stress control. 4. Proposed Research Model\
The present model comprises of four aspects i.e., Work-Life Balance
Variables, Control Variables, Conflict Measurement and Outcome
Variables. The proposed research model depicts the dependency of
work-family conflict and family-work conflict on control variables, the
existing work-life balance variables among doctors and nurses and the
impact of work-life balance variables on job satisfaction, employee
well-being and stress control.

```{r include=FALSE}
data <- read_sav("WLB_final.sav")
#glimpse(data)
library(labelled)
data <-  data %>% unlabelled()
#data11 <- data[,c(17:26,51)]
#data12 <- data[,27:33]
#data13 <- data[,c(34:36,51,50,52)]
#data14 <- data[,c(37,38,45:48)]
#data15 <- data[,c(52:56,39:44)]
#data11[sapply(data11, is.character)] <- lapply (data11[sapply (data11, is.character)], as.factor)
#data11[sapply(data11, is.factor)] <- lapply(data11[sapply(data11, is.factor)], as.ordered)

```

```{r echo=FALSE, fig.height=10, fig.width=10}
data11 <- data[,c(17:26,51)]

#data11[sapply(data11, is.character)] <- lapply(data11[sapply(data11, is.character)], as.factor)
#data11[sapply(data11, is.factor)] <- lapply(data11[sapply(data11, is.factor)], as.ordered)

data11 <-  as.data.frame(data11)
names(data11) = c("Interest Job","Workplace Location","Friendly Environment","Job Security","Training Opportunities","Career development",
"Salary Wages","Working Flexible","Leave","Holidays","Away Home" )

result_wps <- likert(data11, grouping = data$Gender)
plot_wps <- plot(result_wps, centered=TRUE, type = "bar", legend = "") + labs(title = "Work Place Satisfaction", y = "Percent")

plot_wps

# for sector
result_wps_sector<-likert(data11, grouping = data$Working_as)

plot_wps_sector = plot(result_wps_sector , centered=TRUE, type = "bar", legend = "", text.size = 3) + labs(title = "Work Place Satisfaction", x= "Working Sector")
plot_wps_sector

##
data12 <- data[,27:33]
data12 <-  as.data.frame(data12)
names(data12) = c("Ability Untilization","Coworkers","Supervisor","Working Conditions","Medical Facilities","Work Autonomy","Employe Welfare" )

result_wpfs <- likert(data12)
plot_wpfs <- plot(result_wpfs, centered=TRUE, type = "bar", legend = "") + labs(title = "Work Place Facilities Satisfaction", y = "Percent")

plot_wpfs

data13 <- data[,c(34:36,50,52)]
data13 <-  as.data.frame(data13)
names(data13) = c("Emotional Exhaustion","Prolonged Fatigue level","Additional Burdens","Energy","Work Place Problem" )




result_obwp <- likert(data13)
plot_obwp  <- plot(result_obwp , centered=TRUE, type = "bar", legend = "") + labs(title = "Over Burden at Work Place", y = "Percent")
plot_obwp

data14 <- data[,c(37,38,45:48)]
data14 <-  as.data.frame(data14)
names(data14) = c("Family worries","Family Activities","Family Obilgactions","Family Matters","Quality Family Time","Family Holiday" )


result_FE <- likert(data14)
plot_FE   <- plot(result_FE  , centered=TRUE, type = "bar", legend = "") + labs(title = "Family Environment", y = "Percent")
plot_FE

data15 <- data[,c(53:56,39:44)]
data15 <-  as.data.frame(data15)
names(data15) = c("Sense of Pride","Social Status","Policy & Practices","Accomplishment","Self Evelopment","Moral value","Family Commitments","Social Involvement","Quality LIfe","Care Myself" )

result_Ss <- likert(data15)
plot_Ss   <- plot(result_Ss  , centered=TRUE, type = "bar", legend = "") + labs(title = "Social Status", y = "Percent")
plot_Ss
```
